---
title: "Danish Monarchs"
author: "Marie Kern Larsen"
date: " created 4 October 2021, updated `r format(Sys.time(), '%d %B, %Y')` "


output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true

---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the monarchs' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the monarchs

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 


1. Look at the dataset that you are loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator: ; 

2. Create a `monarchs` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# FILL IN THE CODE BELOW and review the outputs
library(tidyverse)

monarchs1 <- read.csv("data/Danish_monarchs.csv")
head(monarchs1)

monarchs2 <- read_csv("data/Danish_monarchs.csv",show_col_types = FALSE)
head(monarchs2)


monarchs3 <- read.csv2("data/Danish_monarchs.csv")
head(monarchs3)

monarchs4 <- read_csv2("data/Danish_monarchs.csv",show_col_types = FALSE)
head(monarchs4) 
```
Answer: 

1. Which of these functions is a `tidyverse` function? Read data with it below into a `monarchs` object

`read_csv()`and `read_csv2()` are both tidyverse functions, I have chosen `read_csv2()`(the same used for monarchs4) because the dataset is danish and seperated with ;

2. What is the result of running `class()` on the `monarchs` object created with a tidyverse function.

The data is loaded as a tibble and therefore the result is "spec_tbl_df" "tbl_df" "tbl" "data.frame"

3. How many columns does the object have when created with these different functions?

The monarchs objects created with `read.csv()`and `read_csv()` only has 1 column, where as the objects created with `read.csv2()`and `read_csv2()` has 9 columns - this is again because it's a danish dataset

4. Show the dataset so that we can see how R interprets each column

```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

monarchs <- read_csv2("data/Danish_monarchs.csv",show_col_types = FALSE)

class(monarchs)


ncol(monarchs1)
ncol(monarchs2)
ncol(monarchs3)
ncol(monarchs4)


glimpse(monarchs)


```


## Calculate the duration of reign for all the monarchs in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the monarchs dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
# YOUR CODE
monarchs %>% 
  mutate(duration=end_of_reign-start_of_reign) %>% 
  select(name_monach,duration)

```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `monarchs` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r}
# YOUR CODE
monarchs %>% 
  mutate(duration=end_of_reign-start_of_reign) %>% 
  summarize(mean_duration=mean(duration,na.rm=TRUE))
```

## How many and which monarchs enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `monarchs` dataset. Display the result and also  count the resulting rows with `count()`

```{r}
# YOUR CODE
monarchs %>% 
  mutate(duration=end_of_reign-start_of_reign) %>%
  select(name_monach, duration) %>% 
  filter(duration>mean(duration,na.rm=TRUE)) %>% 
  count()

monarchs %>% 
  mutate(duration=end_of_reign-start_of_reign) %>%
  select(name_monach, duration) %>% 
  filter(duration>mean(duration,na.rm=TRUE))
  
```


## How many days did the three longest-ruling monarchs rule? 

* Sort monarchs by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
# YOUR CODE
monarchs %>% 
  mutate(duration=end_of_reign-start_of_reign) %>%
  mutate(days=duration*365) %>% 
  select(name_monach, duration,days) %>% 
  arrange(desc(duration)) %>% 
  slice(1:3)
  

```


## Challenge: Plot the monarchs' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the monarchs dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r warning=FALSE}
# YOUR CODE
monarchs_plotting <- monarchs

monarchs_plotting %>%
  mutate(duration=end_of_reign-start_of_reign) %>% 
  mutate(midyear=start_of_reign+(end_of_reign-start_of_reign)/2) %>% 
  ggplot(aes(x=midyear,y=duration))+ geom_point()+geom_smooth()+labs(title="Duration of reign through time",x="Year", y="Duration of reign")+ theme_bw()+theme(text=element_text(size=18))
           
 
```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!